The aims of science and an open science controversy at USEPA

### 2025-05-11 paper/paper.qmd
working on refs

While not as influential as the argument from inductive risk *[Douglas, Havstad]*, the aims approach to values in science provides a distinct challenge to the value-free ideal. Proponents of the approach argue that science typically has not only epistemic aims but also practical or pragmatic ones, and that at any stage of inquiry decisions should be made so as to promote those aims. So, specifically, in the "core" of inquiry decisions should be made so as to promote pragmatic aims, and in this way non-epistemic values can play a legitimate role in the "core" of inquiry. *[cites]*  For example, *[Elliott and McKaughan]* argue that the pragmatic aim of quick and protective regulatory review in chemical safety testing supports using rapid screening techniques, which are much faster than other methods but will likely have a higher false positive rate.

Note that this argument does not necessarily deny that science has epistemic aims, as may be done by approaches inspired by John Dewey *[Anderson; Brown]* or Donald Davidson *[Clough]*. This argument is also independent of questions about the definition of epistemic values or a sharp demarcation of values into "epistemic" vs. "social" *[Longino; Rooney; McMullin; Steel]*. Instead, the key assumption of this argument is that science does not have *only* epistemic aims.

Proponents of the aims approach often argue for pluralism about aims of science *[Elliott and McKaughan; Potochnik]*, with *[Lusk and Elliott]* explicitly rejecting "the view
that science has global aims that constrain the influence of local aims on scientific
practice" (1). Both epistemic and pragmatic aims are often presented as specific to a discipline, field, or research program *[Tapestry ch. 4; examples]*. *[Intemann's]* early statement of the approach argues for "democratically endorsed" aims. *[more?]*

A distinctive aspect of *[Hicks']* version of the aims approach is an analysis of two general "views" of the aims of science, and a claim that an unrecognized disagreement between the two views is a recurrent feature of debates, not only among scientists and the public but also among philosophers.  For Hicks, the "broad view" corresponds to the aims approach argument sketched above — science has both epistemic and pragmatic aims — while the "narrow view" denies the "key assumption" of that argument, taking science to have only epistemic aims *[cites]*. On the narrow view, applications of scientific research are seen as happy accidents but not the primary function or purpose of research. Allowing pragmatic considerations (read: non-epistemic values) to influence inquiry is a corrupting influence, morally on a par with the epistemic distortions introduced by industrial funding *[Bennett and Holman; Fernandez Pinto]* or *[Wansink]*.

@@

*[move most of this to discussion]*
There are also notable differences in the overall tone of the adjectives in supporters' and opponents' top bigrams.  Supporters seem to express skepticism or cynicism about the quality of scientific research: science is described as secret, secretive, phony, false, fake, shady, or deceptive.  In the context of these skeptical assessments, seemingly-positive valence bigrams — such as real, true, verifiable, and sound science — might be used as privatives, that is, expressing the idea that actually-existing science is *not* real, true, verifiable, or sound.  A skeptical or cynical assessment of science would provide support for ST insofar as open data would make science less secretive and phony and more verifiable and true, etc.  @HicksOpenScienceReplication2023 argued that there is no empirical evidence of a replication crisis in environmental epidemiology [though see @BagiletEstimatingSmallEffects *[cite as well in intro]*], and that ST's open data requirements would not mitigate widespread irreplicability even if it were a problem in that field. @HicksDevelopingMeasuresPublic2024 report preliminary evidence that cynicism is a better predictor of generalized trust in science than the value-free ideal.

@@

As an alternative to the aims approach, our findings might be interpreted in terms of inductive risk, and specifically differences between supporters and opponents in the relative importance of over- vs. under-regulation *[cites]*.  That is, supporters of ST might put more weight on over-regulation and preventable economic harms; while opponents of ST might put more weight on under-regulation and preventable disease and loss of life due to pollution.

@@

Second, belief is often understood by philosophers as either a felt commitment to the truth of a claim or proposition, or as a behavioral disposition (with certain cognitive aspects) *[Schwitzgebel]*. In either conceptualization, beliefs are expected to be more-or-less stable across contexts (keeping fixed available evidence, belief in logically relevant claims, etc.). But there is circumstantial evidence^[By this we mean that the evidence below does not show that any identifiable individual endorsed different views of the aims of science in two different contexts. Instead our argument is that two incompatible views appear to be widely endorsed in the same group in different contexts, and thus it is highly probably that many individual members of this group have or would endorse both views in an unstable or logically inconsistent way.] that the views of aims of science we identified here are not stable.

@@

For these reasons, we do not interpret the comments on ST as expressing beliefs of the public on the aims of science. As an alternative, we suggest that the narrow and broad views can be understood as interpretive frames [@KleinDataFrameTheory2007]. Based on studies of the ways professionals make decisions in complex situations (primarily military intelligence analysts), @KleinDataFrameTheory2007 proposed that decisionmakers engage in "sensemaking," a form of inquiry [compare @BrownScienceMoralImagination2020 ch. 1] aimed at understanding rather than empirical accuracy [@KleinDataFrameTheory2007], and specifically understanding to support effective decisionmaking in the face of practical uncertainty. A "frame" is any "explanatory structure[] that account[s] for the data," including but not limited to stories, scripts, and maps [@KleinDataFrameTheory2007 120]. @KleinDataFrameTheory2007 emphasize that there is an iterative relationship between data and frame: "The data identify the relevant frame, and the frame determines which data are noticed. Neither of these comes first" [118; compare *[Hanson, Chang]*]. Because individuals operate with different "repertoires of frames," and data-gathering is informed by frames, "Different people viewing the same events can perceive and recall different things depending on their goals and experiences" [@KleinDataFrameTheory2007 121]. Extending this thought, we suggest that an individual's repertoire might include incompatible frames, and so similar events might be perceived and interpreted in very different ways by the same individual.

Applied to the ST controversy, we suggest that both liberals and conservatives have both narrow and broad views of the aims of science among their repertoire of frames for interpreting public scientific controversies. But the frames were evoked differently in this particular controversy by being connected to two very different frames on the general trustworthiness of EPA/regulatory science *[McCright and Dunlap]*

While the degree of conservative distrust in science is often exaggerated *[cites]*, there is a large difference in partisan attitudes towards USEPA: @CerdaAmericansFeelFavorably2024 reports that only 32% of Republican/Lean Republican survey respondents have a favorable attitude towards USEPA, compared to 73% of Democratic/Lean Democratic respondents. Republican distrust of USEPA appears in the bigram analysis *[xref]*, in the *way* supporters talk about science, using adjectives such as *[list]*. These characterizations of science portray it as suffering from an inappropriate influence of non-epistemic values — specifically, politicization *[cite]*. The narrow view and the value-free ideal justify outrage at this situation, and by extension policies such as ST that are (believed to) mitigate these inappropriate influences. We further suggest that liberal utilization of the narrow view, in the context of scientific integrity plans, is also a response to politicization.

>>>>

While not as influential as the argument from inductive risk [@DouglasInductiveRiskValues2000; @HavstadSensationalScienceArchaic2021], the aims approach to values in science provides a distinct challenge to the value-free ideal. Proponents of the approach argue that science typically has not only epistemic aims but also practical or pragmatic ones, and that at any stage of inquiry decisions should be made so as to promote those aims. So, specifically, in the "core" of inquiry decisions should be made so as to promote pragmatic aims, and in this way non-epistemic values can play a legitimate role in the "core" of inquiry. *[cites]*  For example, @ElliottNonepistemicValuesMultiple2014 argue that the pragmatic aim of quick and protective regulatory review in chemical safety testing supports using rapid screening techniques, which are much faster than other methods but will likely have a higher false positive rate.

Note that this argument does not necessarily deny that science has epistemic aims, as may be done by approaches inspired by John Dewey [@AndersonUsesValueJudgments2004; @BrownScienceMoralImagination2020] or Donald Davidson [@CloughRacistValueJudgments2008; @CloughUsingValuesEvidence2020]. This argument is also independent of questions about the definition of epistemic values or a sharp demarcation of values into "epistemic" vs. "social" [@McMullinValuesScience1982; @RooneyValuesScienceEpistemic1992; @LonginoGenderPoliticsTheoretical1995; @SteelEpistemicValuesArgument2010]. Instead, the key assumption of this argument is that science does not have *only* epistemic aims.

Proponents of the aims approach often argue for pluralism about aims of science [@ElliottNonepistemicValuesMultiple2014; @PotochnikDiverseAimsScience2015; @PotochnikIdealizationAimsScience2017], with @LuskGlobalAimsScience2024 explicitly rejecting "the view that science has global aims that constrain the influence of local aims on scientific practice" (1). Both epistemic and pragmatic aims are often presented as specific to a discipline, field, or research program *[@ElliottTapestryValuesIntroduction2017 ch. 4; examples]*. @IntemannDistinguishingLegitimateIllegitimate2015 coined the term "aims approach," and argued for "democratically endorsed epistemological and social aims of research" (2), which will be pluralistic insofar as different polities democratically endorse different aims. *[more?]*

A distinctive aspect of @HicksWhenVirtuesAre2022 version of the aims approach is an analysis of two general "views" of the aims of science, and a claim that an unrecognized disagreement between the two views is a recurrent feature of debates, not only among scientists and the public but also among philosophers.  For Hicks, the "broad view" corresponds to the aims approach argument sketched above — science has both epistemic and pragmatic aims — while the "narrow view" denies the "key assumption" of that argument, taking science to have only epistemic aims *[cites]*. On the narrow view, applications of scientific research are seen as happy accidents but not the primary function or purpose of research. Allowing pragmatic considerations (read: non-epistemic values) to influence inquiry is a corrupting influence and a form of research misconduct, *morally* on a par with the epistemic distortions introduced by industrial funding [@HolmanExperimentationIndustrialSelection2017; @FernandezPintoEpistemicDiversityIndustrial2023] or fraud [@BrightWhyScientistsLie2021].

@@

@@

As an alternative to the aims approach, our findings might be interpreted in terms of inductive risk, and specifically differences between supporters and opponents in the relative importance of over- vs. under-regulation [compare @HicksScientificControversiesProxy2017 on climate change].  That is, supporters of ST might put more weight on over-regulation and preventable economic harms; while opponents of ST might put more weight on under-regulation and preventable disease and loss of life due to pollution.

@@

Second, belief is often understood by philosophers as either a felt commitment to the truth of a claim or proposition, or as a behavioral disposition (with certain cognitive aspects) [@SchwitzgebelBelief2024]. In either conceptualization, beliefs are expected to be more-or-less stable across contexts (keeping fixed available evidence, belief in logically relevant claims, etc.). But there is circumstantial evidence^[By this we mean that the evidence below does not show that any identifiable individual endorsed different views of the aims of science in two different contexts. Instead our argument is that two incompatible views appear to be widely endorsed in the same group in different contexts, and thus it is highly probably that many individual members of this group have or would endorse both views in an unstable or logically inconsistent way.] that the views of aims of science we identified here are not stable.

@@

For these reasons, we do not interpret the comments on ST as expressing beliefs of the public on the aims of science. As an alternative, we suggest that the narrow and broad views can be understood as interpretive frames [@KleinDataFrameTheory2007]. Based on studies of the ways professionals make decisions in complex situations (primarily military intelligence analysts), @KleinDataFrameTheory2007 proposed that decisionmakers engage in "sensemaking," a form of inquiry [compare @BrownScienceMoralImagination2020 ch. 1] aimed at understanding rather than empirical accuracy [@KleinDataFrameTheory2007], and specifically understanding to support effective decisionmaking in the face of practical uncertainty. A "frame" is any "explanatory structure[] that account[s] for the data," including but not limited to stories, scripts, and maps [@KleinDataFrameTheory2007 120]. @KleinDataFrameTheory2007 emphasize that there is an iterative relationship between data and frame: "The data identify the relevant frame, and the frame determines which data are noticed. Neither of these comes first" [118\; compare @HansonPatternsDiscoveryInquiry1958; @ChangInventingTemperatureMeasurement2004]. Because individuals operate with different "repertoires of frames," and data-gathering is informed by frames, "Different people viewing the same events can perceive and recall different things depending on their goals and experiences" [@KleinDataFrameTheory2007 121]. Extending this thought, we suggest that an individual's repertoire might include incompatible frames, and so similar events might be perceived and interpreted in very different ways by the same individual.

Applied to the ST controversy, we suggest that both liberals and conservatives have both narrow and broad views of the aims of science among their repertoire of frames for interpreting public scientific controversies. But the frames were evoked differently in this particular controversy by being connected to two very different frames on the general trustworthiness of EPA/regulatory science [@McCrightAntireflexivity2010; @McCrightPoliticalPolarizationSupport2014].

While the degree of conservative distrust in science is often exaggerated *[cites]*, there is a large difference in partisan attitudes towards USEPA: @CerdaAmericansFeelFavorably2024 reports that only 32% of Republican/Lean Republican survey respondents have a favorable attitude towards USEPA, compared to 73% of Democratic/Lean Democratic respondents. Republican distrust of USEPA appears in the bigram analysis *[xref]*, in the *way* supporters talk about science. Namely, supporters seem to express skepticism or cynicism about the quality of scientific research: science is described as secret, secretive, phony, false, fake, shady, or deceptive.  In the context of these skeptical assessments, seemingly-positive valence bigrams — such as real, true, verifiable, and sound science — might be used as privatives, that is, expressing the idea that actually-existing science is *not* real, true, verifiable, or sound.  A skeptical or cynical assessment of science would provide support for ST insofar as open data would make science less secretive and phony and more verifiable and true, etc.^[@HicksOpenScienceReplication2023 argued that there is no empirical evidence of a replication crisis in environmental epidemiology [though see @BagiletEstimatingSmallEffects *[cite as well in intro]*], and that ST's open data requirements would not mitigate widespread irreplicability even if it were a problem in that field.] @HicksDevelopingMeasuresPublic2024 report preliminary evidence that cynicism is a better predictor of generalized trust in science than the value-free ideal.

This cynical characterization of science portrays it as suffering from an inappropriate influence of non-epistemic values — specifically, politicization [@BrownScienceDemocracyExpertise2009 ch. 8; @DouglasDifferentiatingScientificInquiry2023]. The narrow view and the value-free ideal justify outrage at this situation, and by extension policies such as ST that are (believed to) mitigate these inappropriate influences. We further suggest that liberal utilization of the narrow view, in the context of scientific integrity plans, is also a response to politicization.
